State-space analysis of continuous gravitational waves with a pulsar timing array: inclusion of the pulsar terms

### 2023-11-28 manuscript/manuscript_2/paper.tex
also do the null model for Bayes

\section{Earth terms vs pulsar terms}\label{sec:earth_vs_psr}
In Section \ref{sec:rep_example1} we have validated the state-space analysis scheme when the pulsar terms are included in the inference model. In this section, critically, we compare the affect of including the pulsar-term in the model (i.e. using Equation \eqref{eq:g_func_trig_chi}), as opposed to the omitting the pulsar-term (i.e. Equation \eqref{eq:g_func_trig_earth}). We do this using the specific representative case of the individual quasi-monochromatic SMBHB source described in Table \ref{tab:parameters_and_priors}. We refer to the model where the pulsar-term is included alongside the Earth term as $\mathcal{M}_{\rm P \& E}$. We refer to the model where the pulsar-term is omitted as 
$\mathcal{M}_{\rm E}$. Model $\mathcal{M}_{\rm P \& E}$ is parameterised by $\boldsymbol{\theta}'$. Model $\mathcal{M}_{\rm E}$ is parameterised by $\boldsymbol{\theta}_{\rm E} = \boldsymbol{\theta}_{\rm gw} \cup \boldsymbol{\theta}'_{\rm psr}$. Both models are applied to identical realisations of the data, generated using the procedure as described in Section  \ref{sec:synthetic_data}. The priors on the static parameters in each model are also identical, with the addition of a uniform prior on $\chi^{(n)}$ for $\mathcal{M}_{\rm P \& E}$, as described in Appendix \ref{sec:set_priors}. \newline

In Section \ref{sec:psr_v_earth_pe} we compare the accuracy of the estimates of $\boldsymbol{\theta'}$ returned by the two models. In Section \ref{sec:psr_v_earth_bayes} we compare the minimal detectable strain for the two models.

@@

We define a relative error metric to quantify the accuracy of the one-dimensional posteriors with respect to the injection value as
\begin{equation}
\Delta_{\rm M}(\theta) = \frac{1}{K}\sum_{k=1}^{K} \frac{p_k(\theta) - \hat{\theta}}{\hat{\theta}} \ , \label{eq:mean_rel_error}
\end{equation}
where $p_k(\theta)$ is the $k$-th one-dimensional posterior sample for parameter $\theta$, $K$ is the total number of samples, $\hat{\theta}$ is the true injection value, and the subscript $M \in \{ \text{earth, pulsar} \}$ indicates whether the posterior is estimated using Equation \eqref{eq:g_func_trig_earth} or Equation \eqref{eq:g_func_trig} respectively. The dependence of $p_k(\theta)$ on $M$ is implied. $\Delta_{\rm M}(\theta)$ represents the mean relative error in the one-dimensional posterior. The values of $\Delta_{\rm M}(\theta)$ for each parameter in Figure \ref{fig:corner_plot_compare_high} are summarised in Table \ref{tab:posterior_errors}. The difference in the mean relative error between the two models, $\Delta_{\rm pulsar}(\theta) - \Delta_{\rm earth}(\theta)$ is also presented. For every parameter except $\Omega$, $\Delta_{\rm pulsar}(\theta) - \Delta_{\rm earth}(\theta)$ is negative, indicating that the parameter estimates are more accurate using the pulsar terms (i.e. the mean relative error when using the pulsar model is smaller). The exception for $\Omega$ is due to the fact that $\Omega$ is recovered with high accuracy in both cases; the the mean relative error is small $\Delta_{\rm M}(\Omega) \sim 10^{-5}$. The improvement from using the pulsar terms is largest for $\iota$, with $\Delta_{\rm pulsar}(\iota) - \Delta_{\rm earth}(\iota) = -0.25$. \newline

For the low-SNR case,  Figure \ref{fig:corner_plot_compare_low}, there is less of a notable improvement in the parameter estimates. Qualitatively, the green and blue one-dimensional posteriors are generally overlaid. The two-dimensional contours suggest a slight improvement through using the pulsar terms, but the difference is modest. The mean relative errors $\Delta_{\rm M}(\theta)$ for the low SNR case are also reported in Table \ref{tab:posterior_errors}. We see that.... \textcolor{red}{TK: numbers to be inserted here}. We conclude that the inclusion of the pulsar terms does not significantly improve the parameter estimates in the low-SNR case. This is intuitive since the system is highly noise dominated such that an increasing model complexity does not aid parameter estimates. Whilst the inclusion of the pulsar terms only modestly improve the accuracy of the parameter estimates in the case of low-SNR, we will show in Section \ref{sec:psr_v_earth_bayes} that they are important for increasing the detection probability for a given signal strength. \newline

@@

\begin{table}
	\centering
	%	\resizebox{\textwidth}{!}{%
		%	\renewcommand{\arraystretch}{1.0} % Default value: 1
		\begin{tabular}{lcclc}
			\toprule
			&$\theta$ & $\Delta_{\rm earth}(\theta)$ & $\Delta_{\rm pulsar}(\theta)$ & $\Delta_{\rm pulsar}(\theta) -\Delta_{\rm earth}(\theta) $   \\
			\hline
			\multirow{7}{2mm}{High SNR} & $\Omega$       & 1.23e-05 &1.49e-05 &2.54e-06 \\
			& $\Phi_0$ &0.0179 &0.0085& -0.00935 \\
			& $\psi$ &0.0386& 0.000254& -0.0384 \\
			& $\iota$ & 0.374 &0.126 &-0.248 \\
			& $\delta$ & 0.00223 &0.000242& -0.00198 \\
			&$\alpha$ &0.0397 &0.000582 &-0.0391 \\
			&$h_0$ & 0.168 &0.0846 &-0.0839 \\
			\hline
			\multirow{7}{2mm}{Low SNR} & $\Omega$       & 1.23e-05 &1.49e-05 &2.54e-06 \\
			& $\Phi_0$ &0.0179 &0.0085& -0.00935 \\
			& $\psi$ &0.0386& 0.000254& -0.0384 \\
			& $\iota$ & 0.374 &0.126 &-0.248 \\
			& $\delta$ & 0.00223 &0.000242& -0.00198 \\
			&$\alpha$ &0.0397 &0.000582 &-0.0391 \\
			&$h_0$ & 0.168 &0.0846 &-0.0839 \\

			\bottomrule
		\end{tabular}
		\caption{Mean relative error $\Delta_{\rm M}(\Theta)$ (Equation \eqref{eq:mean_rel_error}) in the one-dimensional posteriors calculated using the Earth-term model (M = earth, Equation \eqref{eq:g_func_trig_earth}) and the pulsar-term model (M=pulsar, Equation \eqref{eq:g_func_trig}) for the elements $\theta \in \boldsymbol{\theta}_{\rm gw}$. The injection values are as summarised in Table \ref{tab:parameters_and_priors}. The top and bottom halves of the table contain the high-SNR ($h_0 = 10^{-12}$) and low-SNR ($h_0 = 5 \times 10^{-15}$) cases respectively. The rightmost column quotes the difference in the mean relative error between the models for each parameter; a negative value indicates that the pulsar-term model is more accurate.\textcolor{red}{TK numbers to be revised and double checked, but probably representative. Julian notes: swap so positive is good, normalise omega, add units, align}}
		\label{tab:posterior_errors}
	\end{table}

@@

The Bayes factor, $\beta_i$, for each of the two models is presented in Figure \ref{fig:bayes} for the representative source in Table \ref{tab:parameters_and_priors}, except that we now vary the source amplitude, $h_0$, from $10^{-15}$ (undetectable) to $10^{-12}$ (easily detectable). To control the test, the noise processes in the synthetic data are identical realisations for each value of $h_0$; the only change from one $h_0$ value to the next is $h_0$ itself. \footnote{Changing the noise realizations as well, from one value of $h_0$ to the next, adds uninformative scatter to the trend in Figure \ref{fig:bayes}.}

@@

\section{Nested sampling}\label{sec:nested_sampling}
We can use the likelihood returned by the Kalman filter, Equation \eqref{eq:likelihood}, in conjunction with likelihood-based inference methods to estimate the posterior distribution of $\boldsymbol{\theta}$ by Bayes' Rule,
\begin{equation}
	p(\boldsymbol{\theta} | \boldsymbol{Y}) = \frac{\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta}) \pi(\boldsymbol{\theta})}{\mathcal{Z}} \ ,
\end{equation}
where $\pi(\boldsymbol{\theta})$ is the prior distribution on $\boldsymbol{\theta}$ and $\mathcal{Z}$ is the marginalised likelihood, or evidence
\begin{equation}
	\mathcal{Z} = \int d \boldsymbol{\theta} \mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta})  \pi(\boldsymbol{\theta})  \ . \label{eq:model_evidence}
\end{equation}
We estimate the posterior distribution and the model evidence through nested sampling \citep{Skilling} in this paper. Nested sampling evaluates marginalised likelihood integrals, of the form given by Equation \eqref{eq:model_evidence}. It also approximates the posterior by returning samples from $p(\boldsymbol{\theta} | \boldsymbol{Y})$. It does so by drawing a set of $n_{\rm live}$ live points from $\pi(\boldsymbol{\theta})$ and iteratively replacing the live point with the lowest likelihood with a new live point drawn from $\pi(\boldsymbol{\theta})$, where the new live point is required to have a higher likelihood than the discarded point. The primary advantage of nested sampling is its ability to compute $\mathcal{Z}$, on which model selection relies. Nested sampling is also computationally efficient and can handle multi-modal problems \citep{Ashton2022}. For these reasons, it has enjoyed widespread adoption in the physical sciences, particularly within the cosmological community \citep{Mukherjee2006,Feroz2008,Handley2015}, neutron star astrophysics \citep{Myers2021MNRAS.502.3113M,Meyers2021,Melatos2023}, particle physics \citep{proceedings2019033014} and materials science \citep{2009arXiv0906materials}. For reviews of nested sampling we refer the reader to \cite{Buchner2021} and \cite{Ashton2022}. Multiple nested sampling algorithms and computational libraries exist. \citep[e.g.][]{Feroz2008,Feroz2009,Handley2015,dynesty2020,UltraNest2021}. In gravitational wave research it is common to use the \texttt{dynesty} sampler \citep{dynesty2020} via the \texttt{Bilby} \citep{bilby.507.2037A} front-end library. We follow this precedent and use \texttt{Bilby} for all nested sampling Bayesian inference in this work. \newline

@@

For $\pi(\boldsymbol{\theta}_{\rm gw})$ we choose standard non-informative priors \citep[e.g.][]{Bhagwat2021}. For $\pi(\boldsymbol{\theta}_{\rm psr})$ \textit{a priori} information from electromagnetic observations does exist. We adopt constrained uniform priors on $f_{\rm em}^{(n)}(t_1)$ and $\dot{f}_{\rm em}^{(n)}(t_1)$, which extend $\pm 10^3 \eta_f^{(n)}$ and $\pm 10^3 \eta_{\dot{f}}^{(n)}$ respectively about the central, injected values, where $\eta_f^{(n)}$ and $\eta_{\dot{f}}^{(n)}$ denote the errors quoted in the ATNF Pulsar Database. By using wider-than-necessary priors we expose the analysis scheme to a more stringent test. We set an uninformative broad prior $\pi[\sigma^{(n)} / (1 \, {\rm s^{-3/2}})] \sim$ LogUniform($10^{-23}, 10^{-19}$) which covers the range of values inferred from Equation \eqref{eq:sigmap_f}. We set an non-informative prior on the new static parameter $\chi^{(n)} \sim \text{Uniform}\left(0,2\pi\right)$. We do not set a prior on $\gamma^{(n)}$, because $\gamma^{(n)} T_{\rm obs} \sim 10^{-5}$, and consequently $\gamma^{(n)}$ is effectively ``unobservable'' over $T_{\rm obs}$. For validation purposes it is therefore sufficient to carry $\gamma^{(n)}$ through the analysis at its injected value. This reduces the total dimension of the parameter space to $7 + 4N$. The chosen prior $\boldsymbol{\theta}$ is the same as used in K24, with the addition of a prior on $\chi^{(n)}$ due to the inclusion of the pulsar term. \newline

The chosen priors for the representative analysis in Section \ref{sec:rep_example1} are summarised in \ref{Table tab:parameters_and_priors}

\section{Complete workflow summary}

For the reader's convenience we now summarise the workflow for a representative PTA analysis using the Kalman filter and nested sampler for parameter estimation and model selection:
\begin{enumerate}[leftmargin=2em]
	\item Specify a PTA composed of $N$ pulsars 
	\item Obtain $N$ data inputs $f_{\rm m}^{(n)}(t)$, collectively labelled $\boldsymbol{Y}$
	\item Specify a state-space model $\mathcal{M}$, with static parameters $\boldsymbol{\theta}$
	\item Specify prior distribution $\pi(\boldsymbol{\theta})$
	\item Sample $n_{\rm live}$ points from $\pi(\boldsymbol{\theta})$ 
	\item For each live point:
	\begin{enumerate}[leftmargin=2em]
		\item Pass the sample $\boldsymbol{\theta}_{\rm sample}$ to the Kalman filter
		\item Iterate over the input data using the Kalman filter and obtain a single $\log \mathcal{L}$ value, Equation \eqref{eq:likelihood}
	\end{enumerate}
	\item Remove the live point with the lowest likelihood value, $\log \mathcal{L}_{\rm lowest}$
	\item Sample a new live point from $\pi(\boldsymbol{\theta})$, subject to the requirement that the new likelihood obeys $\mathcal{L}_{\rm new}$ > $\mathcal{L}_{\rm lowest}$, where $\log \mathcal{L}_{\rm new}$ is calculated via steps (vi)(a)--(vi)(b).
	\item Update $p\left(\boldsymbol{\theta}|\boldsymbol{Y}\right)$ and $\mathcal{Z}$ with nested sampler
	\item Repeat steps (vii)--(ix) until convergence criteria are satisfied.
\end{enumerate}
In order to compute $\beta$ the above workflow is repeated for a different $\mathcal{M}$. The resulting $\mathcal{Z}$ values can then be compared. We remind the reader that the above workflow differs from a realistic PTA analysis in one important respect, namely that the data are input as frequency time series $f_{\rm m}^{(n)}(t)$ instead of pulse TOAs. The generalization to TOAs is subtle and will be tackled in a forthcoming paper.

\section{Canonical example, high SNR} \label{sec:app_high_SNR}

		\begin{figure}
	\includegraphics[width=\columnwidth, height = \columnwidth ]{example-image-a} 	
	\caption{}
	\label{fig:corner_high_snr_appendix}
\end{figure}

	\begin{figure}
	\includegraphics[width=\columnwidth, height = \columnwidth ]{example-image-b} 	
	\caption{}
	\label{fig:corner_high_snr_appendix_chi}
\end{figure}

\section{Multiple noise realisaitons and the Wasserstein distance}

We can extend the analysis of these two points and consider 1000 noise realisations of $\boldsymbol{Y}$, beyond the \textcolor{red}{X} plotted in Figure \ref{fig:corner_plot_1}. As in Figure \ref{fig:corner_plot_1} we independently estimate the static parameters, $\boldsymbol{\theta}$ for each realisation and calculate $p({\boldsymbol{\theta'}} | {\boldsymbol{Y}})$.

To this end we start with the representative example in Table \ref{tab:parameters_and_priors} and generate $1000$ realisations of the process noise $\xi^{(n)}(t)$ and measurement noise $\varepsilon^{(n)}(t)$. For each realisation we

We now consider the complete set of $10^3$ noise realisations, going beyond the subset of nine realisations in the preceding discussion. We ask the question: how ``similar'' are the $10^3$ marginalized, one-dimensional posteriors computed for each of the seven parameters in ${\boldsymbol{\theta}}_{\rm gw}$? There is no unique way to answer this question. In this paper we appeal to the Wasserstein distance \citep[WD;][]{Wasserstein,Villani2009} from optimal transport theory, which defines an intuitive notion of similarity between probability distributions. The WD is a popular metric in machine learning \citep{2017arXiv170107875A}, climate modelling \citep{2022JCli...35.1215P,2023QJRMS.149..843K}, computational biology \citep{GONZALEZDELGADO2023168053} and geophysics \citep{2023GeoRL..5003880M}; a short review is presented in Appendix \ref{sec:wasserstein}. The WD measures the cost of an optimal strategy for moving probability mass between two distributions from position $x$ to position $y$, with respect to some cost function $c(x,y)$. In this paper we use the first WD moment, $W_1(\mu,\nu)$, which is defined and interpreted in Appendix \ref{sec:wasserstein}. Intuitively $W_1$ bounds the difference in the expectation value of a parameter selected from ${\boldsymbol{\theta}}_{\rm gw}$ with respect to the PDFs $\mu$ and $\nu$. Taking a concrete example, suppose that we infer two one-dimensional posterior distributions $\mu(\iota)$ and $\nu(\iota)$ for $\iota$, for two different realisations of the noise, and calculate $W_1(\mu, \nu) =0.5$ rad. Then we can infer $| \langle \iota \rangle_\mu - \langle \iota \rangle_\nu | \leq 0.5 \, {\rm rad}$. \newline

Table \ref{tab:Wasserstein} summarizes the WD between the $5\times 10^5$ pairs of one-dimensional posteriors across the $10^3$ realizations, for each of the seven parameters in ${\boldsymbol{\theta}}_{\rm gw}$. The median $W_1$ for each parameter is tabulated in the rightmost column and denoted by $W_{1,{\rm median}}$. As a percentage of the known, injected value, $W_{\rm 1,median}$ ranges from  0.2\% for $\Omega$ to 67\% for $\Phi_0$. These values quantify approximately the natural dispersion in parameter estimates when the analysis scheme is applied to real, astronomical data, where the true parameter values are unknown. Moreover, as a percentage of the prior domain, $W_{\rm 1,median}$ ranges from $8.1 \times 10^{-4}$ \% for $h_0$ to 2.7\% for $\psi$. These values confirm that the nested sampler converges reliably to a single, narrow peak without railing against the prior bounds for $10^3$ noise realizations. Further analysis of the WD is performed in Appendix \ref{sec:wasserstein}.

>>>>

\section{Earth terms vs pulsar terms}\label{sec:earth_vs_psr}
In Section \ref{sec:rep_example1} we have validated the state-space analysis scheme when the pulsar terms are included in the inference model. In this section, critically, we compare the effect of including the pulsar-term in the model (i.e. using Equation \eqref{eq:g_func_trig_chi}), as opposed to the omitting the pulsar-term (i.e. Equation \eqref{eq:g_func_trig_earth}). We do this using the specific representative case of the individual quasi-monochromatic SMBHB source described in Table \ref{tab:parameters_and_priors}. We refer to the model where the pulsar-term is included alongside the Earth term as $\mathcal{M}_{\rm P \& E}$. We refer to the model where the pulsar-term is omitted as 
$\mathcal{M}_{\rm E}$. Model $\mathcal{M}_{\rm P \& E}$ is parameterised by $\boldsymbol{\theta}'$. Model $\mathcal{M}_{\rm E}$ is parameterised by $\boldsymbol{\theta}_{\rm E} = \boldsymbol{\theta}_{\rm gw} \cup \boldsymbol{\theta}'_{\rm psr}$. Both models are applied to identical realisations of the data, generated using the procedure as described in Section  \ref{sec:synthetic_data}. The priors on the static parameters in each model are also identical, with the addition of a uniform prior on $\chi^{(n)}$ for $\mathcal{M}_{\rm P \& E}$, as described in Appendix \ref{sec:set_priors}. \newline

In Section \ref{sec:psr_v_earth_pe} we compare the accuracy of the estimates of $\boldsymbol{\theta'}$, $\boldsymbol{\theta}_{\rm E}$ returned by the respective models. In Section \ref{sec:psr_v_earth_bayes} we compare the minimal detectable strain for the two models.

@@

We define a root-mean-square relative error to quantify the accuracy of the one-dimensional posteriors with respect to the injection value as
\begin{equation}
\Delta_{\rm M}(\theta) = \sqrt{\frac{1}{K}\sum_{k=1}^{K} \frac{p_k(\theta) - \hat{\theta}}{\hat{\theta}}} \ , \label{eq:mean_rel_error}
\end{equation}
where $p_k(\theta)$ is the $k$-th one-dimensional posterior sample for parameter $\theta$, $K$ is the total number of samples, $\hat{\theta}$ is the true injection value, and the subscript $M \in \{ \text{earth, pulsar} \}$ indicates whether the posterior is estimated using Equation \eqref{eq:g_func_trig_earth} or Equation \eqref{eq:g_func_trig} respectively. The dependence of $p_k(\theta)$ on $M$ is implied. The values of $\Delta_{\rm M}(\theta)$ for each parameter in Figure \ref{fig:corner_plot_compare_high} are summarised in Table \ref{tab:posterior_errors}. The difference in the error between the two models, $\Delta_{\rm earth}(\theta)-\Delta_{\rm pulsar}(\theta)$ is also presented. For every parameter $\Omega$, $\Delta_{\rm earth}(\theta)-\Delta_{\rm pulsar}(\theta)$ is positive, indicating that the parameter estimates are more accurate using the pulsar terms. In some cases the difference is modest;  $\Omega$ is recovered with high accuracy by both models such that $\Omega$, $\Delta_{\rm earth}(\Omega)-\Delta_{\rm pulsar}(\Omega) = 6.7 \times 10^{-3}$. The improvement from using the pulsar terms is largest for $\iota$, with $\Delta_{\rm earth}(\iota) - \Delta_{\rm pulsar}(\iota)  = 0.24$. \newline

For the low-SNR case,  Figure \ref{fig:corner_plot_compare_low}, there is less of a notable improvement in the parameter estimates. Qualitatively, the green and blue one-dimensional posteriors are generally overlaid. The two-dimensional contours suggest a slight improvement through using the pulsar terms, but the difference is modest. The root-mean-square relative error $\Delta_{\rm M}(\theta)$ for the low SNR case is also reported in Table \ref{tab:posterior_errors}. We see that the error is larger for every element in $\theta_{\rm gw}$ than in the high SNR case, as expected.

The inclusion of the pulsar terms improve the estimates for five out of the seven static parameters; $\Delta_{\rm earth}(\theta)-\Delta_{\rm pulsar}(\theta)$ is positive for all parameters except $\psi$ and $h_0$. However it is hard to draw strong conclusions; for different realisations of the noise different parameters exhibit either improvement or deterioration in $Delta_{\rm M}(\theta)$. We conclude that the inclusion of the pulsar terms does not significantly improve the parameter estimates in the low-SNR case. This is intuitive since the system is highly noise dominated such that an increasing model complexity does not aid parameter estimates. Whilst the inclusion of the pulsar terms only modestly improve the accuracy of the parameter estimates in the case of low-SNR, we will show in Section \ref{sec:psr_v_earth_bayes} that they are important for increasing the detection probability for a given signal strength. \newline

@@

\begin{table}
	\centering
	%	\resizebox{\textwidth}{!}{%
		%	\renewcommand{\arraystretch}{1.0} % Default value: 1
		\begin{tabular}{lcllc}
			\toprule
			&$\theta$ & $\Delta_{\rm earth}(\theta)$ & $\Delta_{\rm pulsar}(\theta)$ & $\Delta_{\rm earth}(\theta) - \Delta_{\rm pulsar}(\theta)$   \\
			\hline
			\multirow{7}{2mm}{High SNR} & $\Omega$       & 0.011& 0.0043& 0.00671\\
			& $\Phi_0$ &0.00332 &0.00115 &0.00217\\
			& $\psi$ &0.0968 &0.000653 &0.0962 \\
			& $\iota$ & 0.343 &0.106 &0.237 \\
			& $\delta$ & 0.00187 &0.000285 &0.00158 \\
			&$\alpha$ &0.0403 &0.000435 &0.0399 \\
			&$h_0$ & 0.227 &0.075 &0.152 \\
			\hline
			\multirow{7}{2mm}{Low SNR} & $\Omega$       & 2.7 &1.6& 1.09 \\
			& $\Phi_0$ &0.335 &0.229 &0.106 \\
			& $\psi$ &0.112 &0.123 &-0.0119 \\
			& $\iota$ & 0.443 &0.37 &0.0724 \\
			& $\delta$ & 0.181 &0.0782 &0.103 \\
			&$\alpha$ &0.145 &0.139 &0.00637 \\
			&$h_0$ & 1.12 &1.41 &-0.291 \\

			\bottomrule
		\end{tabular}
		\caption{Root-mean-square relative error $\Delta_{\rm M}(\theta)$ ,Equation \eqref{eq:mean_rel_error}, in the one-dimensional posteriors calculated using the Earth-term model (M = earth, Equation \eqref{eq:g_func_trig_earth}) and the pulsar-term model (M=pulsar, Equation \eqref{eq:g_func_trig}) for the elements $\theta \in \boldsymbol{\theta}_{\rm gw}$. The injection values are as summarised in Table \ref{tab:parameters_and_priors}. The top and bottom halves of the table contain the high-SNR ($h_0 = 10^{-12}$) and low-SNR ($h_0 = 5 \times 10^{-15}$) cases respectively.  For all parameters the units are in rad, except for $h_0$ which is dimensionless. The rightmost column quotes the difference in the mean relative error between the models for each parameter; a negative value indicates that the pulsar-term model is more accurate.}
		\label{tab:posterior_errors}
	\end{table}

@@

The Bayes factor, $\beta_i$, for each of the two models is presented in Figure \ref{fig:bayes1} for the representative source in Table \ref{tab:parameters_and_priors}, except that we now vary the source amplitude, $h_0$, from $10^{-15}$ (undetectable) to $10^{-12}$ (easily detectable).

For a given $h_0$, we generate 10 realisations of synthetic data

To control the test, the noise processes in the synthetic data are identical realisations for each value of $h_0$; the only change from one $h_0$ value to the next is $h_0$ itself. \footnote{Changing the noise realizations as well, from one value of $h_0$ to the next, adds uninformative scatter to the trend in Figure \ref{fig:bayes}.}

@@

\section{Nested sampling}\label{sec:nested_sampling}
We can use the likelihood returned by the Kalman filter, Equation \eqref{eq:likelihood}, in conjunction with likelihood-based inference methods to estimate the posterior distribution of $\boldsymbol{\theta}$ by Bayes' Rule,
\begin{equation}
	p(\boldsymbol{\theta} | \boldsymbol{Y}) = \frac{\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta}) \pi(\boldsymbol{\theta})}{\mathcal{Z}} \ ,
\end{equation}
where $\pi(\boldsymbol{\theta})$ is the prior distribution on $\boldsymbol{\theta}$ and $\mathcal{Z}$ is the marginalised likelihood, or evidence
\begin{equation}
	\mathcal{Z} = \int d \boldsymbol{\theta} \mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta})  \pi(\boldsymbol{\theta})  \ . \label{eq:model_evidence2}
\end{equation}
We estimate the posterior distribution and the model evidence through nested sampling \citep{Skilling} in this paper. Nested sampling evaluates marginalised likelihood integrals, of the form given by Equation \eqref{eq:model_evidence2}. It also approximates the posterior by returning samples from $p(\boldsymbol{\theta} | \boldsymbol{Y})$. It does so by drawing a set of $n_{\rm live}$ live points from $\pi(\boldsymbol{\theta})$ and iteratively replacing the live point with the lowest likelihood with a new live point drawn from $\pi(\boldsymbol{\theta})$, where the new live point is required to have a higher likelihood than the discarded point. The primary advantage of nested sampling is its ability to compute $\mathcal{Z}$, on which model selection relies. Nested sampling is also computationally efficient and can handle multi-modal problems \citep{Ashton2022}. For these reasons, it has enjoyed widespread adoption in the physical sciences, particularly within the cosmological community \citep{Mukherjee2006,Feroz2008,Handley2015}, neutron star astrophysics \citep{Myers2021MNRAS.502.3113M,Meyers2021,Melatos2023}, particle physics \citep{proceedings2019033014} and materials science \citep{2009arXiv0906materials}. For reviews of nested sampling we refer the reader to \cite{Buchner2021} and \cite{Ashton2022}. Multiple nested sampling algorithms and computational libraries exist. \citep[e.g.][]{Feroz2008,Feroz2009,Handley2015,dynesty2020,UltraNest2021}. In gravitational wave research it is common to use the \texttt{dynesty} sampler \citep{dynesty2020} via the \texttt{Bilby} \citep{bilby.507.2037A} front-end library. We follow this precedent and use \texttt{Bilby} for all nested sampling Bayesian inference in this work. \newline

@@

For $\pi(\boldsymbol{\theta}_{\rm gw})$ we have no \textit{a priori} information on the parameters. We therefore choose standard non-informative priors \citep[e.g.][]{Bhagwat2021}. For $\pi(\boldsymbol{\theta}_{\rm psr})$ \textit{a priori} information from electromagnetic observations does exist. We adopt constrained uniform priors on $f_{\rm em}^{(n)}(t_1)$ and $\dot{f}_{\rm em}^{(n)}(t_1)$, which extend $\pm 10^3 \eta_f^{(n)}$ and $\pm 10^3 \eta_{\dot{f}}^{(n)}$ respectively about the central, injected values, where $\eta_f^{(n)}$ and $\eta_{\dot{f}}^{(n)}$ denote the errors quoted in the ATNF Pulsar Database. By using wider-than-necessary priors we expose the analysis scheme to a more stringent test. We set an uninformative broad prior $\pi[\sigma^{(n)} / (1 \, {\rm s^{-3/2}})] \sim$ LogUniform$ \left [10^{-2} \sigma_{\rm SC}^{(n)}, 10^2 \sigma_{\rm SC}^{(n)} \right ]$
where $\sigma_{\rm SC}^{(n)}$ is the value for pulsar $n$ inferred from
Equation \eqref{eq:sigmap_f}. We set an non-informative prior on the new static parameter $\chi^{(n)} \sim \text{Uniform}\left(0,2\pi\right)$. We do not set a prior on $\gamma^{(n)}$, because $\gamma^{(n)} T_{\rm obs} \sim 10^{-5}$, and consequently $\gamma^{(n)}$ is effectively ``unobservable'' over $T_{\rm obs}$. For validation purposes it is therefore sufficient to carry $\gamma^{(n)}$ through the analysis at its injected value. This reduces the total dimension of the parameter space to $7 + 4N$. \newline

\section{Complete workflow summary}
For the reader's convenience we now summarise the workflow for a representative PTA analysis using the Kalman filter and nested sampler for parameter estimation and model selection:
\begin{enumerate}[leftmargin=2em]
	\item Specify a PTA composed of $N$ pulsars 
	\item Obtain $N$ data inputs $f_{\rm m}^{(n)}(t)$, collectively labelled $\boldsymbol{Y}$
	\item Specify a state-space model $\mathcal{M}$, with static parameters $\boldsymbol{\theta}$
	\item Specify prior distribution $\pi(\boldsymbol{\theta})$
	\item Sample $n_{\rm live}$ points from $\pi(\boldsymbol{\theta})$ 
	\item For each live point:
	\begin{enumerate}[leftmargin=2em]
		\item Pass the sample $\boldsymbol{\theta}_{\rm sample}$ to the Kalman filter
		\item Iterate over the input data using the Kalman filter and obtain a single $\log \mathcal{L}$ value, Equation \eqref{eq:likelihood}
	\end{enumerate}
	\item Remove the live point with the lowest likelihood value, $\log \mathcal{L}_{\rm lowest}$
	\item Sample a new live point from $\pi(\boldsymbol{\theta})$, subject to the requirement that the new likelihood obeys $\mathcal{L}_{\rm new}$ > $\mathcal{L}_{\rm lowest}$, where $\log \mathcal{L}_{\rm new}$ is calculated via steps (vi)(a)--(vi)(b).
	\item Update $p\left(\boldsymbol{\theta}|\boldsymbol{Y}\right)$ and $\mathcal{Z}$ with nested sampler
	\item Repeat steps (vii)--(ix) until convergence criteria are satisfied.
\end{enumerate}
In order to compute $\beta$ the above workflow is repeated for a different $\mathcal{M}$. The resulting $\mathcal{Z}$ values can then be compared. We remind the reader that the above workflow differs from a realistic PTA analysis in one important respect, namely that the data are input as frequency time series $f_{\rm m}^{(n)}(t)$ instead of pulse TOAs. The generalization to TOAs is subtle and will be tackled in a forthcoming paper.

\section{Parameter estimation of representative example, high-SNR} \label{sec:app_high_SNR}

In this Section we repeat the analysis of Section \ref{sec:pe_and_ms}

In this section, we apply the Kalman filter in conjunction with nested sampling to calculate the joint posterior probability distribution $p({\boldsymbol{\theta'}} | {\boldsymbol{Y}})$ for synthetic data $\boldsymbol{Y}$, and compare it to the known, injected values. The procedure is undertaken for multiple realisations of the noise processes $\xi^{(n)}(t)$ and $\varepsilon^{(n)}(t)$. That is, given the injected static parameters $\boldsymbol{\theta'}$, multiple realisations of $\boldsymbol{Y}$ are generated using the methods described in Section \ref{sec:synthetic_data}, and for each realisation we independently estimate $\boldsymbol{\theta'}$. The aims are (i) to demonstrate that the analysis scheme works (i.e.\ that it converges to a well-behaved, unimodal posterior for multiple noise realisations), (ii) to give a preliminary sense of its accuracy, and (iii) quantify the natural random dispersion in the one-dimensional posterior medians between noise realisations. 
Regarding point (iii), quantifying the dispersion is important since it is a practical measure of the accuracy of the parameter estimation scheme when it is applied to real astronomical data, where the true parameter values and specific noise realisations are unknown. The injected static GW source parameters $\boldsymbol{\theta}_{\rm gw}$ used for this mock analysis to generate $\boldsymbol{Y}$ are selected to be astrophysically reasonable and representative. Different values of the static GW source parameters are explored in Section \ref{sec:parameter_space}. The static pulsar parameters $\boldsymbol{\theta'}_{\rm psr}$  are as specified in Section \ref{sec:synthetic_data}. All injected static parameters $\boldsymbol{\theta}'$ are summarised in the second column of Table \ref{tab:parameters_and_priors}. The specification of the priors on $\boldsymbol{\theta'}$ is described in Appendix \ref{sec:set_priors} and summarised in the rightmost column of Table \ref{tab:parameters_and_priors}. \newline

		\begin{figure*}
	\includegraphics[width=\textwidth, height = \textwidth ]{images/corner_highSNR} 	
	\caption{}
	\label{fig:corner_high_snr_appendix}
\end{figure*}

	\begin{figure}
	\includegraphics[width=\columnwidth, height = \columnwidth ]{images/corner_highSNR_chi} 	
	\caption{}
	\label{fig:corner_high_snr_appendix_chi}
\end{figure}

%
%\section{Multiple noise realisaitons and the Wasserstein distance}
%
%
%We can extend the analysis of these two points and consider 1000 noise realisations of $\boldsymbol{Y}$, beyond the \textcolor{red}{X} plotted in Figure \ref{fig:corner_plot_1}. As in Figure \ref{fig:corner_plot_1} we independently estimate the static parameters, $\boldsymbol{\theta}$ for each realisation and calculate $p({\boldsymbol{\theta'}} | {\boldsymbol{Y}})$. 
%
%
%
%
%To this end we start with the representative example in Table \ref{tab:parameters_and_priors} and generate $1000$ realisations of the process noise $\xi^{(n)}(t)$ and measurement noise $\varepsilon^{(n)}(t)$. For each realisation we 
%
%
